Supplementary Materials

### 2023-07-25 manuscript/manuscript.tex
ms updates from italy, this time with covid

%\title{Towards a Complete Model of the Negative Selection Process in Humans}
%\title{An Evolutionary Quantitative Genetic Model of the Negative Selection Process in Humans}
\title{A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans}

@@

\subsection*{Predicted Substitution Rate and Additive Fitness Variance}

@@

    \caption{Predicted versus observed divergence along the human lineage.
        Black points are the PhyloFit divergence rate estimates per features
        (on x-axis). Line ranges are the implied divergences across a range of
        human-chimpanzee divergence times of 6-12 Mya (using a generate time of
        29 years). We show the predicted divergences for our Feature
        (turquoise) and PhastCons priority (green) annotation models.
        Additionally, we show the predicted PhastCons region divergences when
    local-rescaling is applied (we omit other locally-rescaled predictions
since these to not differ substantially).}

@@

There are five important but tentative results to draw from this analysis.
First, the estimated mutation rate is implausibly higher, reaching the upper
limit of our optimization boundary of $\mu = 8 \times 10^{-8}$ across all
populations in or PhastCons Priority model (Supplementary Material
\ref{supp:locally-rescaled-fits}). When we re-ran optimization with an upper
limit of $\mu = 1 \times 10^{-7}$, the mutation rate was again estimated to
this (implausibly large) upper limit, with no considerable changes to model fit
or DFE estimates. Second, all of our leave-one-chromosome-out $R_\text{LOCO}^2$
are about one percentage point higher than the unrescaled model. Third, the DFE
estimates for PhastCons regions is now U-shaped (Supplementary Materials Figure
\ref{suppfig:rescale-dfe}), with $75-77\%$ of mass being placed on a weakly
deleterious class, $s=10{^-5}$. Interestingly, this is the first of all of our
models where such an appreciable mass has been placed on a midpoint in our
grid; in all other cases, non-strongly estimates were neutral ($s=10^{-8}$).
Fourth, predicted pairwise diversity is nearly identical to our original,
non-locally-rescaled fit (see Supplementary Materials Figure
\ref{suppfig:rescale}). Finally, since local-rescaling increases the DFE mass
over $s < 10^{-3}$, mutations in PhastCons regions now have the possibility of
fixation. We find that local-rescaling the PhastCons Priority, leads the
predicted substitution rates in PhastCons regions to be much closer to observed
levels (blue line, Figure \ref{fig:figure-6}).

Finally, we one important caveat about this analysis. Since local-rescaling is
done using the first round of maximum likelihood estimates, there is a
possibility of some circularity, since the $B(x)$ at this position includes the
contribution of the focal segment that is being rescaled. Ideally, one would
exclude this segment's contribution to $B(x)$; however, this is computationally
unfeasible. However, two observations indicate our findings here are robust
despite this limitation. First, the results do not change based on whether the
$B(x)$ is averaged at the 1 kbp level or at the megabase scale; for the latter,
a single segment makes little contribution to the average. Second, we
investigated the extent to which local-rescaling differed from the B' maps
across selection parameters. We find minor differences between the
locally-rescaled and standard B' maps in the nearly-neutral domain ($0.2 \le
2N_e s \le 2$); the maps were identical under strong selection ($2N_e s= 20$)
as expected (Supplementary Materials Figure \ref{suppfig:rescale-b}). Moreover,
the correlations between the standard and local-rescaled B' maps across the
genome are high (100\% for $2N_e s = 20$, 96.5\% for $2N_e s = 2$, and 60.21\%
for $2N_e s = 0.2$). The overall realized effect of local-rescaling is to just
alter how deep the ``U" is in the relationship between the reduction factor $B$
and the selection coefficient (Supplementary Materials Figure
\ref{suppfig:rescale-effects}).

Overall, this suggests that models that fit the linked selection are worryingly
sensitive the theoretic B' values in the $2Ns \le 1$ domain. The fact that
predicted diversity differs little between standard and locally-rescaled B'
methods indicates there may not be enough information in the linked selection
signal alone to differentiate when interference is occurring or the causes of
fitness variance along the genome. Moreover, local-rescaling turns out to only
slightly alter the B' maps, yet significantly modifies the DFE estimates. This
brings the deleterious substitution rate in agreement with observations (since
this is predicted with a fixed $\mu = 1.5 \times 10^{-8}$; however, the maximum
likelihood estimate of mutation is implausibly high.

@@

Since the SC16 theory calculates fitness variance by jointly predicting the
deleterious substitution rate, this provides another route to verify our main
findings. We find that predicted substitution rates for highly-conserved
PhastCons features are in strong disagreement with observed substitution rates
along the human lineage: the DFE-substitution rate contradiction. This is a
consequence of our DFE estimate for PhastCons regions containing little mass
over weakly deleterious and neutral selection coefficients --- a characteristic
of DFE estimates from other work too \parencite{Murphy2022-sj}. One clue to
this discrepancy is that our simulation results indicate an appreciable level
of error between observed reductions and the B' maps in the weak selection
domain of $2N_e s \approx 1$.

We hypothesized that this could be because in this domain, a segment under
selective constraint experiences a local draft-effective population size of
$B(x)N$, and not just $N$. This local draft-effective population size is
induced by selection at other segments that is not being taken into account by
classic BGS theory or our standard model. When we experimented by fitting our
model and then using the predicted reduction map to local-rescale $N_e$ to the
draft-effective population size $N_d = \widehat{B}(x)$, we find the
disagreement between predicted and observed substitution rates disappears. This
is expected, since locally-rescaled fits result in DFE estimates to have an
appreciable mass on weakly deleterious selection coefficients. While this
brings substitution rates into agreement, we encounter a similar problem found
by \textcite{McVicker2009-ax}: the estimated mutation rate is too high. We
emphasize that although there are several caveats with the local-rescaling
approach, in practice it is altering the B' maps only slightly in the domain of
$2N_e s \approx 1$ (deepening the bottom of the ``U"). Even if the local-rescaling
approximation to selective interference is incorrect, these results suggest
parameter estimates are extremely sensitivity to how accurate the B' maps are
in this domain.

Unfortunately, this work opens up more questions about the nature of selection
in humans that it answers. The implausibly high estimate of mutation rates
could suggest that negative selection is not the only source of fitness
variance generating the genome-wide linked selection signal in humans. Positive
selection on polygenic variation along the genome could be another source of
fitness variance contributing to the weak selection signal. Alternatively,
negative selection could be the main source of fitness variance, but the
complexity of selective interference may be poorly approximated by the
local-rescaling approach. Another possibility is that our assumption throughout
of additive effects may lead to biases given deleterious mutations likely
recessive effects. Overall, understanding how the sources of human fitness
variance shape the striking patterns of diversity along the human genome still
remains an open riddle.

@@

where,

\begin{align}
  %V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  V_a &= (U - 2 R) s_j & \text{\emph{fitness variance equation}} \\
  Q^2(m_i, s_j) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us_j}{U-{M}} & \text{\emph{variance decay rate}}
\end{align}
%
and $U = m_i L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates in the segment. A detailed derivation of these equations
can be found in Supplementary Materials Section \ref{supp:theory}. The recombination
rate in a segment is determined by a user-supplied recombination map.

@@

Following previous approaches
\parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}, we use a composite
likelihood approach to fit our negative selection model. Per-basepair allele
count data (described below) is summarized into the number of same and
different pairwise differences per window. All of our primary models were fit
with megabase windows, since previous work has found the strongest selection
signal at this scale (we confirm this with one XXX fit at the $10^5$bp scale).

@@

Our method uses two tricks to improve optimization over the mutation and DFE
parameters. First, we find $B(x; \mu, \mathbf{W})$ is exponential over columns
of $\mu \mathbf{W}$, which allows for optimization over this smooth function
rather than the grid. Second, we use softmax to convert constrained
optimization over the DFE columns (which must sum to one) to unconstrained. We
tested multiple different optimization routines, finding that BOBYQA
outperformed alternatives \parencite{Powell2009-jm,Johnson2007-tl}. We
inspected and confirmed convergence with diagnostic plots finding stable optima
across 10,000 random starts (see Supplementary Materials Section
\ref{supp:optima}). We assessed model fit using out-sample predictive error,
calculated by leaving out a whole chromosome during fitting and predicting its
diversity. To calculate uncertainty, we used a block jackknife approach in 10
Mbp windows (Supplementary Materials Section \ref{supp:jackknife}).

@@

>>>>

%\title{Towards a Complete Model of the Negative Selection Process in Humans}
%\title{An Evolutionary Quantitative Genetic Model of the Negative Selection Process in Humans}
\title{A Quantitative Genetic Linked Selection Model of Linked Selection in Humans}

@@

\subsection*{Predicted Substitution Rates}

@@

    \caption{Predicted versus observed divergence along the human lineage.
        Black points are the PhyloFit divergence rate estimates per features
        (on x-axis). Line ranges are the implied divergences across a range of
        human-chimpanzee divergence times of 6-12 Mya (using a generate time of
        29 years). We show the predicted divergences for our Feature
        (turquoise) and PhastCons priority (green) annotation models.
        Additionally, we show the predicted PhastCons region divergences when
    local-rescaling is applied (blue; we omit other locally-rescaled predictions
since these to not differ substantially).}

@@

There are five important but tentative results to draw from this analysis.
First, estimated mutation rates are in general higher. Under the CADD 6\%
model, they are $\widehat{\mu} = 6-7.3 \times 10^{-8}$ across populations; for
the PhastCons Priority model, they reach the upper limit of our optimization
boundary of $\mu = 8 \times 10^{-8}$ (Supplementary Material
\ref{supp:locally-rescaled-fits}). Second, all of our leave-one-chromosome-out
$R_\text{LOCO}^2$ are about one percentage point higher than the unrescaled
model. Third, the DFE estimates for both CADD 6\% and PhastCons regions in the
PhastCons Priority model is now U-shaped (Supplementary Materials Figure
\ref{suppfig:rescale-dfe}), with $70-77\%$ of mass being placed on a weakly
deleterious class, $s=10{^-5}$. Interestingly, this is the first of all of our
models where such an appreciable mass has been placed on a midpoint in our
selection coefficient grid; in all other cases, non-strongly estimates were
neutral ($s=10^{-8}$). Fourth, predicted pairwise diversity is nearly identical
to our original, non-locally-rescaled fit (see Supplementary Materials Figure
\ref{suppfig:rescale}). Finally, since local-rescaling increases the DFE mass
over $s < 10^{-3}$, mutations in PhastCons regions now have the possibility of
fixation. We find that local-rescaling the PhastCons Priority, leads the
predicted substitution rates in PhastCons regions to be much closer to observed
levels (blue line, Figure \ref{fig:figure-6}).

Finally, we note an important caveat about this analysis. Since local-rescaling
is done using the first round of maximum likelihood estimates, there is a
possibility of some circularity, since the $B(x)$ at this position includes the
contribution of the focal segment that is being rescaled. Ideally, one would
exclude this segment's contribution to $B(x)$; however, this is computationally
unfeasible. However, two observations indicate our findings here are relatively
robust despite this limitation. First, the results do not change based on
whether the $B(x)$ is averaged at the 1 kbp level or at the megabase scale; for
the latter, a single segment makes little contribution to the average. Second,
we investigated the extent to which local-rescaling modified the B' maps across
selection parameters. We find minor differences between the locally-rescaled
and standard B' maps in the nearly-neutral domain ($0.2 \le 2N_e s \le 2$); the
maps were identical under strong selection ($2N_e s= 20$) as expected
(Supplementary Materials Figure \ref{suppfig:rescale-b}). Moreover, the
correlations between the standard and local-rescaled B' maps across the genome
are high (100\% for $2N_e s = 20$, 96.5\% for $2N_e s = 2$, and 60.21\% for
$2N_e s = 0.2$). The overall realized effect of local-rescaling is to just
alter how deep the ``U" is in the relationship between the reduction factor $B$
and the selection coefficient (Supplementary Materials Figure
\ref{suppfig:rescale-effects}).

Overall, this suggests that models that fit the linked selection are worryingly
sensitive to the theoretic B' values in the $2Ns \le 1$ domain. The fact that
predicted diversity differs little between standard and locally-rescaled B'
methods indicates there may not be enough information in the linked selection
signal alone to differentiate when interference is occurring or the causes of
fitness variance along the genome. Moreover, local-rescaling turns out to only
slightly alter the B' maps, yet significantly modifies the DFE estimates. This
brings the deleterious substitution rate in agreement with observations (since
this is predicted with a fixed $\mu = 1.5 \times 10^{-8}$; however, the maximum
likelihood estimate of mutation is implausibly high. This suggests either the
local-rescaling approximation to interference is not suitable (though
chromosome-wide simulations show it results in lower error), or there is
another selection process contributing to total fitness variance.

@@

Since the SC16 theory calculates fitness variance by jointly predicting the
deleterious substitution rate, this provides another avenue to verify or refute
our models. We find that predicted substitution rates for highly-conserved
PhastCons features are in strong disagreement with observed substitution rates
along the human lineage: the DFE-substitution rate contradiction. This is a
consequence of our DFE estimate for PhastCons regions containing little mass
over weakly deleterious and neutral selection coefficients --- a characteristic
of DFE estimates from other work too \parencite{Murphy2022-sj}. One clue to
this discrepancy is that our simulation results indicate an appreciable level
of error between observed reductions and the B' maps in the weak selection
domain of $2N_e s \approx 1$.

We hypothesized that this could be because in this domain, a segment under
selective constraint experiences a local draft-effective population size of
$B(x)N$, and not just $N$. This local draft-effective population size is
induced by selection at other segments that is not being taken into account by
classic BGS theory or our standard model. When we experimented by fitting our
model and then using the predicted reduction map to local-rescale $N_e$ to the
draft-effective population size $N_d = \widehat{B}(x)$, we find the
disagreement between predicted and observed substitution rates disappears. This
is expected, since locally-rescaled fits result in DFE estimates to have an
appreciable mass on weakly deleterious selection coefficients. While this
brings substitution rates into agreement, we encounter a similar problem found
by \textcite{McVicker2009-ax}: the estimated mutation rate is too high. We
emphasize that although there are several caveats with the local-rescaling
approach, in practice it is altering the B' maps only slightly in the domain of
$2N_e s \approx 1$ (deepening the bottom of the ``U"). Even if the
local-rescaling approximation to selective interference is incorrect (though
simulations indicate it does reduce error), these results suggest parameter
estimates are extremely sensitivity to how accurate the B' maps are in this
domain. Moreover, we find that predicted diversity differs little across DFE
and mutation rate estimates, suggesting there may be limited information in the
pairwise diversity linked selection signal to differentiate models. While
nucleotide diversity is a less informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. While one could imagine using more informative SFS-based
models to infer selection parameters, accounting for selective interference in
these models would be difficult.

Unfortunately, this work opens up more questions about the nature of selection
in humans that it answers. The high estimate of mutation rates could suggest
that negative selection is not the only source of fitness variance generating
the genome-wide linked selection signal in humans. Positive selection on
polygenic variation along the genome could be another source of fitness
variance contributing to the weak selection signal. Alternatively, negative
selection could be the main source of fitness variance, but the complexity of
selective interference may be poorly approximated by the local-rescaling
approach. Another possibility is that our assumption throughout of additive
effects may lead to biases given deleterious mutations likely recessive
effects. As we find evidence of strong selection against loss-of-function in
our model residuals, it is also possible that the bulk of fitness variance
\emph{is} due to purifying selection, but our model is unable to account for
local strong effects since the DFE is estimated per feature. Overall,
understanding how the sources of human fitness variance shape the striking
patterns of diversity along the human genome still remains an open riddle.

\subsection*{Acknowledgments}

We would like to thank Nate Pope and Rasmus Nielsen for helpful advice, and
Martin Kircher for providing special CADD tracks. This research was supported
by NIH awards R35GM148253 and R01HG010774 to ADK

@@

where,
%
\begin{align}
  %V(x) &= U s - 2 {R}(x) s & \text{\emph{fitness variance equation}} \\
  V_a &= (U - 2 R) s_j & \text{\emph{fitness variance equation}} \\
  Q^2(m_i, s_j) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage inflation factor}} \\
  Z &= 1 - \frac{Us_j}{U-{M}} & \text{\emph{variance decay rate}}
\end{align}
%
and $U = m_i L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates in the segment. A detailed derivation of these equations
can be found in Supplementary Materials Section \ref{supp:theory}. The recombination
rate in a segment is determined by a user-supplied recombination map.

@@

Following previous approaches
\parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}, we use a composite
likelihood approach to fit our negative selection model. Per-basepair allele
count data (described below) is summarized into the number of same and
different pairwise differences per window. All of our primary models were fit
with megabase windows, since previous work has found the strongest selection
signal at this scale (we confirm this with one CADD 6\% fit at the $100$ kbp
scale).

@@

Our method uses two tricks to improve optimization over the mutation and DFE
parameters. First, we find $B(x; \mu, \mathbf{W})$ is exponential over columns
of $\mu \mathbf{W}$, which allows for optimization over this smooth function
rather than the grid. Second, we use softmax to convert constrained
optimization over the DFE columns (which must sum to one) to unconstrained. We
tested multiple different optimization routines, finding that BOBYQA
outperformed alternatives \parencite{Powell2009-jm,Johnson2007-tl}. We
inspected and confirmed convergence with diagnostic plots finding stable optima
across 10,000 random starts (see Supplementary Materials Section
\ref{supp:optim}). We assessed model fit using out-sample predictive error,
calculated by leaving out a whole chromosome during fitting and predicting its
diversity. To calculate uncertainty, we used a block jackknife approach in 10
Mbp windows (Supplementary Materials Section \ref{supp:jackknife}).

@@

\subsection*{Data Availability}

All code from \texttt{bgspy} and our Jupyter Lab \parencite{Kluyver_undated-ol}
notebooks for analysis are available on GitHub
(\url{https://github.com/vsbuffalo/bprime}). The main model fits are available
as Python Pickle objects on FigShare (XXX).

### 2023-07-25 manuscript/supplementary.tex
ms updates from italy, this time with covid

Linked selection theory typically assumes that focal sites are not under
purifying selection themselves. Consequently, linked selection statistical
methods often only consider the fraction of the genome not \emph{a priori}
under selective constraint \parencite[e.g. Appendix Section
3.1]{Murphy2022-sj}. Our methods follow suit, and first create a track
consisting of the union (i.e. \texttt{bedtools merge}) of: exons and coding
sequences from Ensembl Version 107
\parencite[\path{Homo_sapiens.GRCh38.107.chr.gff3.gz}]{Cunningham2022-vk} and
PhastCons regions
\parencite[\path{phastConsElements100way.bed.gz}]{Siepel2005-wh}. Then, we
remove 200 basepairs of ``slop" around each of these regions (i.e.
\texttt{bedtools slop}). We show in Supplementary Figure
\ref{suppfig:neutral-tracks} that average diversity (we only consider YRI
samples for this analysis) is fairly insensitive to the slop width, but
increasing slop widths around putatively conserved segments does lead to higher
diversity estimates. This is expected, since increasing the slop width
considers a smaller subset of sites, further away from putatively conserved
regions. For our analyses we choose the 200bp slop version, which covers
63.66\% of the autosomal reference sequence and consists of 1.26 million
contiguous genomic regions.

@@

\clearpage
\subsection{Locally-Rescaled Fit Results}

@@

\clearpage
\subsubsection{Locally-Rescaled Model Fit Results}
\label{supp:locally-rescaled-fits}

@@

>>>>

Linked selection theory typically assumes that focal sites are not under
purifying selection themselves. Consequently, linked selection statistical
methods often only consider the fraction of the genome not \emph{a priori}
under selective constraint \parencite[e.g. Appendix Section
3.1]{Murphy2022-sj}. Our methods follow suit, and first create a track
consisting of the union (i.e. \texttt{bedtools merge}) of: exons and coding
sequences from Ensembl Version 107
\path{Homo_sapiens.GRCh38.107.chr.gff3.gz}]{Cunningham2022-vk} and PhastCons
regions \path{phastConsElements100way.bed.gz}]{Siepel2005-wh}. Then, we remove
200 basepairs of ``slop" around each of these regions (i.e. \texttt{bedtools
slop}). We show in Supplementary Figure \ref{suppfig:neutral-tracks} that
average diversity (we only consider YRI samples for this analysis) is fairly
insensitive to the slop width, but increasing slop widths around putatively
conserved segments does lead to higher diversity estimates. This is expected,
since increasing the slop width considers a smaller subset of sites, further
away from putatively conserved regions. For our analyses we choose the 200bp
slop version, which covers 63.66\% of the autosomal reference sequence and
consists of 1.26 million contiguous genomic regions.

@@

\clearpage
\subsection{Locally-Rescaled Fit Results}
\label{supp:locally-rescaled-fits}

@@

\subsubsection{PhastCons Priority Locally-Rescaled Model Fit Results}

@@

\subsubsection{CADD 6\% Locally-Rescaled Model Fit Results}

\begin{minipage}{\linewidth}\begin{footnotesize}
\begin{verbatim}
SimplexModel (interpolated w): 6 x 7 x 1
  w grid: [1.000e-11 6.310e-11 3.981e-10 2.512e-09 1.585e-08 1.000e-07] (before interpolation)
  t grid: [1.e-08 1.e-07 1.e-06 1.e-05 1.e-04 1.e-03 1.e-02]
  window size: 1.0 Mbp

ML estimates: (whole genome)
standard error method: None
negative log-likelihood: 61820130669.36475
number of successful starts: 10000 (100.0% total)
pi_0 = 0.00102026
pi  = 0.000752601
mut. rate  = 7.291e-08
Ne = 25,506 (if μ=1e-8), Ne = 12,753 (if μ=2e-8)
R² = 60.6057% (in-sample)
W =
          cadd6
------  -------
1e-08     0.009
1e-07     0
1e-06     0
1e-05     0.749
0.0001    0.014
0.001     0.067
0.01      0.161
\end{verbatim}
\end{footnotesize}\end{minipage}

\begin{minipage}{\linewidth}\begin{footnotesize}
\begin{verbatim}
SimplexModel (interpolated w): 6 x 7 x 1
  w grid: [1.000e-11 6.310e-11 3.981e-10 2.512e-09 1.585e-08 1.000e-07] (before interpolation)
  t grid: [1.e-08 1.e-07 1.e-06 1.e-05 1.e-04 1.e-03 1.e-02]
  window size: 1.0 Mbp

ML estimates: (whole genome)
standard error method: None
negative log-likelihood: 197825902519.15137
number of successful starts: 10000 (100.0% total)
pi_0 = 0.00107426
pi0  = 0.000801495
mut. rate  = 5.991e-08
Ne = 26,856 (if μ=1e-8), Ne = 13,428 (if μ=2e-8)
R² = 62.0668% (in-sample)
W =
          cadd6
------  -------
1e-08     0.001
1e-07     0
1e-06     0
1e-05     0.71
0.0001    0.027
0.001     0.065
0.01      0.197
\end{verbatim}
\end{footnotesize}\end{minipage}

\begin{minipage}{\linewidth}\begin{footnotesize}
\begin{verbatim}
SimplexModel (interpolated w): 6 x 7 x 1
  w grid: [1.000e-11 6.310e-11 3.981e-10 2.512e-09 1.585e-08 1.000e-07] (before interpolation)
  t grid: [1.e-08 1.e-07 1.e-06 1.e-05 1.e-04 1.e-03 1.e-02]
  window size: 1.0 Mbp

ML estimates: (whole genome)
standard error method: None
negative log-likelihood: 250648005868.94275
number of successful starts: 10000 (100.0% total)
pi0 = 0.0014366
pi  = 0.00106447
mut. rate  = 6.024e-08
Ne = 35,914 (if μ=1e-8), Ne = 17,957 (if μ=2e-8)
R² = 68.6747% (in-sample)
W =
          cadd6
------  -------
1e-08     0
1e-07     0
1e-06     0
1e-05     0.707
0.0001    0.048
0.001     0.012
0.01      0.233
\end{verbatim}
\end{footnotesize}\end{minipage}
